How to \emph{not} write Virtual Machines for Dynamic Languages

### 2014-09-11 talk/dyla2007/dyla.tex
work on the proposal

PyPy contains a Python interpreter implemented in Python, from which
Python VMs can be generated.  The reader is referred to
\cite{pypyvmconstruction} for a technical presentation.  Let us emphasis
that the argument we make in the present paper is \emph{not} that
VMs for dynamic languages should be written in their own host language
(as many projects like Squeak \cite{Squeak} and others have done) but
instead that VMs should not be \emph{written} in the first place -- they
should be generated from simple interpreters written in any suitable
high-level\footnote{``High-level'' is taken by opposition to languages
like Scheme48's PreScheme \cite{kelsey-prescheme} or Squeak's \cite{Squeak}
SLang, which use the syntax and
metaprogramming facilities of a high-level language but encode
low-level details like object layout and memory management.} language.

In section \ref{sect:approaches} we will explore the way VMs are typically
implemented in C and on top of OO VMs and some of the problems of these
approaches, using various Python implementations as the main example. In
section \ref{sect:metaprogramming} we will describe our proposed
meta-programming approach and compare the two solutions.  We summarize
our position and conclude in section \ref{sect:conclusion}.

@@

\item
\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a
higher memory overhead to start with.  For example, an instance of Sun's
JVM which just loaded Jython consumes between 34 and 42 MB of memory
(Linux/IA32), while on the same machine a CPython process fits in 3 to 4
MB.

@@

\section{Meta-Programming Is Good}
\label{sect:metaprogramming}

@@

\subsection{PyPy architecture}

We implemented this idea in the PyPy project \cite{pypy}.  The dynamic language
for which we wrote an interpreter is Python.  It is a language which,
because of its size and rather intricate semantics, is a good target for
our approach in the following sense: its previous reimplementations
(Jython for the JVM and IronPython for .NET) have each proved to be very
time-consuming to maintain.  Our implementation is, by construction,
easier to maintain and extremely portable (including to C/POSIX, to the
JVM and to .NET).

@@

Writing a good garbage collector remains hard, though.  At least, it is
easy to experiment with various kind of GCs, so we started by just using
the conservative Boehm \cite{Boehm} collector for C and moved up to a
range of simple custom collectors -- reference counting, mark-and-sweep,
etc.  Ultimately, though, more advanced GCs will be needed to get the
best performance.  It seems that RPython, enhanced with support for
direct address manipulations, is a good language for writing GCs, so it
would be possible for a GC expert to write one for our translation
framework.  However, this is not the only way to obtain good GCs:
existing GCs can also be reused.  Good candidates are the GCs written in
the Jikes RVM \cite{JikesGC}.  As they are written in Java, it should be
relatively straightforward to design a translation step that turns them
into RPython (or directly into our RPython-level intermediate
representation).  Our translation toolchain could then integrate any one
of these GCs into the VMs it generates.

@@

\begin{itemize}
\item \emph{Let's write more meta-programming translation toolchains.}
Aside from the advantages described in section
\ref{sect:metaprogramming}, a translation toolchain need not be
standardized for inter-operability but can be tailored to the needs of
each project.
\item \emph{Diversity is good.}  Meta-programming translation
toolchains offset the need for standardization of general-purpose OO VMs.
\end{itemize}

>>>>

PyPy contains a Python interpreter implemented in Python, from which
Python VMs can be generated.  The reader is referred to
\cite{pypyvmconstruction} for a technical presentation.  Let us emphasis
that the argument we make in the present paper is \emph{not} that
VMs for dynamic languages should be written in their own host language
(as many projects like Squeak \cite{Squeak} and others have done) but
instead that VMs should not be \emph{written} in the first place -- they
should be generated from simple interpreters written in any suitable
high-level\footnote{``High-level'' is taken by opposition to languages
like Scheme48's PreScheme \cite{kelsey-prescheme} or Squeak's \cite{Squeak}
SLang, which use the syntax and
meta-programming facilities of a high-level language but encode
low-level details like object layout and memory management.} language.

In section \ref{sect:approaches} we will explore the way VMs are typically
implemented in C and on top of OO VMs and some of the problems of these
approaches, using various Python implementations as the main example. In
section \ref{sect:meta-programming} we will describe our proposed
meta-programming approach and compare the two solutions.  We summarize
our position and conclude in section \ref{sect:conclusion}.

@@

\item
\emph{Better GCs:} While this is obvious in theory, OO VMs tend to have a
higher memory overhead to start with.  For example, an instance of Sun's
JVM which just loaded Jython consumes between 34 and 42 MB of memory
(Linux/IA32), while on the same machine a CPython process fits into 3 to 4
MB.

@@

\section{Meta-Programming Is Good}
\label{sect:meta-programming}

@@

\subsection{PyPy's architecture}

We implemented this idea in the PyPy project \cite{pypy}.  The dynamic language
for which we wrote an interpreter is Python.  It is a language which,
because of its size and rather intricate semantics, is a good target for
our approach in the following sense: its previous re-implementations
(Jython for the JVM and IronPython for .NET) have each proved to be very
time-consuming to maintain.  Our implementation is, by construction,
easier to maintain and extremely portable (including to C/POSIX, to the
JVM and to .NET).

@@

Writing a good garbage collector remains hard, though.  At least, it is
easy to experiment with various kind of GCs, so we started by just using
the conservative Boehm \cite{Boehm} collector for C and moved up to a
range of simple custom collectors -- reference counting, mark-and-sweep,
etc.  Ultimately, though, more advanced GCs will be needed to get the
best performance.  It seems that RPython, enhanced with support for
direct address manipulations, is a good language for writing GCs, so it
would be possible for a GC expert to write one for our translation
framework.  However, this is not the only way to obtain good GCs:
existing GCs could also be reused.  Good candidates are the GCs written in
the Jikes RVM \cite{JikesGC}.  As they are written in Java, it should be
relatively straightforward to design a translation step that turns them
into RPython (or directly into our RPython-level intermediate
representation).  Our translation toolchain could then integrate any one
of these GCs into the VMs it generates.

@@

\begin{itemize}
\item \emph{Let's write more meta-programming translation toolchains.}
Aside from the advantages described in section
\ref{sect:meta-programming}, a translation toolchain need not be
standardized for inter-operability but can be tailored to the needs of
each project.
\item \emph{Diversity is good.}  Meta-programming translation
toolchains offset the need for standardization of general-purpose OO VMs.
\end{itemize}
